Strong lensing selection effects

### 2022-08-19 paper/seleff.tex
paragraph on lensed quasars

We describe the population of background quasars with the following double power-law luminosity function in the rest-frame UV absolute magnitude $M$:
\begin{equation}
\Phi(M,\zqso) = \dfrac{\Phi(M^*)}{10^{0.4(\alpha_Q+1)(M - M^*)} + 10^{0.4(\beta_Q+1)(M - M^*)}}.
\end{equation}
Following \citet{Man++17}, we set $\alpha_Q=-1.35$ (faint-end slope), $\beta_Q=-3.23$ (bright-end slope), and adopted a redshift-evolving normalisation
\begin{equation}
\log{\Phi^*} = -6.0991 + 0.0209\zqso + 0.0171\zqso^2,
\end{equation}
and charachteristic magnitude
\begin{equation}
M^* = -22.5216 - 1.6510\zqso + 0.2869\zqso^2.
\end{equation}
Given the redshift and rest-frame UV luminosity of a quasar, we then computed the apparent magnitude in the observed $i-$band, $\mqso$, using a quasar spectral template\footprint{\url{https://archive.stsci.edu/hlsps/reference-atlases/cdbs/grid/comp_qso/}} built from optical and near-infrared spectra obtained by \citet{vdB++01,Gli++06}.

@@

\Fref{fig:teinhist} shows the distribution in Einstein radius of the simulated lens samples.
In all cases, the distribution peaks at $\tein\approx0.7''$.
Although all of the lenses in these samples are detected, it does not necessarily follow that they would all be included in a strong lensing study.
There can be a few reasons for escluding certain lenses from a sample. 
One is the low accuracy of lens finders: current automated lens finding algorithms tend to produce lens candidates samples with low purity \citep[see e.g.][]{Son++18a,Pet++19,Sav++22}. Such samples are then visually inspected, and only those candidates that can be clearly distinguished from false-positives are kept. This visual inspection step tends to disfavour lenses with a small image separation, because of the contamination from the light of the lens galaxy.
\begin{figure}
\includegraphics[width=\columnwidth]{tein_hist.eps}
\caption{
Einstein radius distribution of the simulated lens samples.
\label{fig:teinhist}
}
\end{figure}

Another reason for refining a sample of lens candidates is the availability of redshift measurements. Redshifts of both the lens and the source are needed in order to convert a lens model into a measurement of mass. When working with large samples of lenses, obtaining spectroscopic measurements is not a viable option, and photometric redshifts are a necessity. Measuring the photometric redshift of a strongly lensed source, however, is challenging, especially when the Einstein radius is small and the source light is blended with the light from the lens \citep{Lan++22}.

@@

Qualitatively, the results shown in \Fref{fig:lenspars} and \Fref{fig:sourcepars} match our expectations from \Sref{sect:indlenses}.
In the rest of this section we quantify the lensing bias.
We present the results in three different parts.
First, we focus on the properties of the lens galaxies that can be observed directly. These are quantities that can be derived from photometry and spectroscopy with minimal assumptions: the lens redshift, the observed stellar mass, the half-light radius and the axis ratio.
Any bias in these quantities can be determined relatively easily in a real strong lens survey.
In the second part, we consider the parameters related to the mass distribution: the stellar population synthesis mismatch parameter and the dark matter distribution parameters. Determining the bias on these parameters is much more difficult, but these are quantities of great interest from a galaxy science point of view.
In the third part, we focus on source distribution parameters.

@@

>>>>

We describe the population of background quasars with the following double power-law luminosity function in the rest-frame UV absolute magnitude $M$:
\begin{equation}
\Phi(M,\zqso) = \dfrac{\Phi(M^*)}{10^{0.4(\alpha_Q+1)(M - M^*)} + 10^{0.4(\beta_Q+1)(M - M^*)}}.
\end{equation}
Following \citet{Man++17}, we set $\alpha_Q=-1.35$ (faint-end slope), $\beta_Q=-3.23$ (bright-end slope), and adopted a redshift-evolving normalisation
\begin{equation}
\log{\Phi^*} = -6.0991 + 0.0209\zqso + 0.0171\zqso^2,
\end{equation}
and charachteristic magnitude
\begin{equation}
M^* = -22.5216 - 1.6510\zqso + 0.2869\zqso^2.
\end{equation}
Given the redshift and rest-frame UV luminosity of a quasar, we then computed the apparent magnitude in the observed $i-$band, $\mqso$, using a quasar spectral template\footnote{\url{https://archive.stsci.edu/hlsps/reference-atlases/cdbs/grid/comp_qso/}} built from optical and near-infrared spectra obtained by \citet{vdB++01,Gli++06}.

@@

\Fref{fig:teinhist} shows the distribution in Einstein radius of the simulated lens samples.
In all cases, the distribution peaks at $\tein\approx0.7''$.
Although all of the lenses in these samples are detected, it does not necessarily follow that they would all be included in a strong lensing study.
There can be a few reasons for escluding certain lenses from a sample. 
One is the low accuracy of lens finders: current automated lens finding algorithms tend to produce lens candidates samples with low purity \citep[see e.g.][]{Son++18a,Pet++19,Sav++22}. Such samples are then visually inspected, and only those candidates that can be clearly distinguished from false-positives are kept. This visual inspection step tends to disfavour lenses with a small image separation, because of the contamination from the light of the lens galaxy.
\begin{figure}
\includegraphics[width=\columnwidth]{tein_hist.eps}
\caption{
Einstein radius distribution of the simulated lens samples.
These are: galaxy-galaxy lenses in the fiducial, high scatter and low scatter scenarios; galaxy-quasar lenses with fiducial scatter, considering all lenses or only lenses that produce four images.
\label{fig:teinhist}
}
\end{figure}

Another possible reason for refining a sample of lens candidates is the availability of redshift measurements. Redshifts of both the lens and the source are needed in order to convert a lens model into a measurement of mass. When working with large samples of lenses, obtaining spectroscopic measurements is not a viable option, and photometric redshifts are a necessity. Measuring the photometric redshift of a strongly lensed source, however, is challenging, especially when the Einstein radius is small and the source light is blended with the light from the lens \citep{Lan++22}.

@@

Qualitatively, the results shown in \Fref{fig:lenspars} and \Fref{fig:sourcepars} match our expectations from \Sref{sect:indlenses}.
In the rest of this section we quantify the lensing bias.
We present the results in three different parts.
First, we focus on the properties of the lens galaxies that can be observed directly. These are quantities that can be derived from photometry and spectroscopy with minimal assumptions: the lens redshift, the observed stellar mass, the half-light radius and the axis ratio\footnote{Strictly speaking, the axis ratio in mass is not necessarily observable, but in the context of our simulations it is, since light and mass have the same ellipticity.}.
Any bias in these quantities can be determined relatively easily in a real strong lens survey.
In the second part, we consider the parameters related to the mass distribution: the stellar population synthesis mismatch parameter and the dark matter distribution parameters. Determining the bias on these parameters is much more difficult, but these are quantities of great interest from a galaxy science point of view.
In the third part, we focus on source distribution parameters.

@@

The main goal of the experiment with lensed quasars is to check whether there are any differences in the strong lensing bias with respect to the extended source case, at fixed properties of the foreground galaxy population.
For this reason, we only ran simulations with lensed quasars in the fiducial scatter scenario and compared the results with those from the extended source simulation.
%We are also interested in understanding how the entire population of quasar lenses differs from the subset of quad lenses, therefore we computed the bias for 
We are also interested in understanding how the subset of quad lenses differs from the entire population of quasar lenses, therefore we computed the bias for this sample as well.
%\Fref{fig:quasarbias} shows the strong lensing bias obtained with the lensed quasars simulation, compared with that 
\Fref{fig:quasarbias} shows the strong lensing bias as a function of minimum Einstein radius in the following quantities: stellar population synthesis mismatch parameter, halo mass at fixed stellar mass, enclosed dark matter mass at fixed stellar mass and half-light radius, and axis ratio.

\begin{figure}
\includegraphics[width=\columnwidth]{quasar_bias.eps}
\caption{

}
\end{figure}
